---
title: "Bat Star Larvae Settlement"
format: html
editor: visual
author: Alondra Casas
---

### STAT 210 Portfolio

## Part 1 - Describe your Process

# Data Description

This is a data set of larval settlement of bat stars (*Patiria miniata*) over time. This data set is from my personal research project that I've been doing over the past few months. My proposed research question is "How does the substrate type affect the rate of *Patiria miniata* larvae settlement over time? This data set is a table of variables such as date, counts of settled stars and larvae, the container numbers and the corresponding treatments they're in. This data comes from my personal project so I'm the source and some cleaning I would like to do is filter for when larvae is greater than one, mutate the date column, add an "experiment date" column to show trend over days, and maybe adding proportions.

# Data Cleaning

During the data cleaning process I filtered for when the larvae is greater than one because I didn't want any zeros in my visualization skewing the plots. Then I added a column called "Status" to show the exact counts of larvae and settled stars for each container each day. I then added a column to calculate proportion to see the rate of settlement over time.

You can add options to executable code like this

```{r}
#| echo: false
library("tidyverse")
library(ggplot2)
```

```{r}
star <- read_csv("data-raw/Casas Bat star Project Data - Sheet1.csv")
```

```{r}
#filtering for when larvae is at least 1 because I want to try running statistics and it doesn't like zeros, also might just make the graph look better
star_clean <- star |>
  filter(Larvae >= 1)

#making a column for status to label each container for each day with the exact counts of larvae and settled stars
star_clean2 <- star_clean |>
    pivot_longer(cols = c(Larvae, Settled), 
               names_to = "Status", 
               values_to = "Count")

#adding proportion to show a more distinct change in settlement rate
starclean3 <- star_clean |>
  mutate( Total = Larvae + Settled,
          propsettled = (Settled / Total)* 100)


```

```{r}
write_csv(star_clean, "data-clean/star_clean.csv")
write_csv(star_clean2, "data-clean/star_clean2.csv")
write_csv(starclean3, "data-clean/starclean3.csv")

```

# Visualization 1 - Settlement Over Time 

This is my first visualization which is a scatter plot showing the number of settled stars over time (in days) for each treatment. Each color represents a treatment which is labeled in the legend. As you can see from the graph, the sea grape treatments had a large impact on settlement.

```{r}
#| fig-alt: A scatterplot graph showing the number of settled stars over time in days for each substrate treatment. Each day is plotted with a dot and a smoothed line was added to show the linear regression over time. The treatments are labeled on the legend (control, rhodomonas only, sterilized sea grapes, and non-sterilized sea grapes). The sea grape treatments show a much faster settlement rate over the experiment.
star |>
  ggplot(aes(x = Day, y = Settled, color = Treatment)) +
  geom_point(size = 0.5) +
  geom_smooth() +
  scale_x_continuous(breaks = 1:27) +
  labs(title = "Number of settled larvae over time", 
       x = "Days",
       y = "Number of settled larvae",
       color = "Treatment") +
  theme_bw() +
  theme(legend.position = "bottom") +
  theme(legend.text = element_text(size = 5))
```

Figure 1: This is a scatter plot showing the number of settled stars throughout the experiment in days. Each treatment is represented by a different color which are labeled by the legend .

# Visualization 2 - Proportion of Settled Stars

This is a box plot showing the proportion of settled stars in each treatment at the end of the experiment. Proportion is on the y-axis and is shown on a scale from 0-100. We see a much higher proportion of settled stars in the non-sterilized sea grape treatment.Each color is representative of a treatment which is labeled in the legend (control, rhodomonas only, sterilized sea grapes, and non-sterilized sea grapes).

```{r}
#| fig-alt: A boxplot showing the proportion of settled stars in each treatment at the end of the experiment. Legend corresponds to the treatments(control, rhodomonas only, sterilized sea grapes, and non-sterilized sea grapes). Non-sterilized sea grapes had a much higher proportion of settled stars compared to the other groups.
starclean3 |>
  ggplot(aes(y = propsettled, color = Treatment, fill = Treatment, )) +
  geom_boxplot(alpha = 0.5) +
  scale_x_continuous(breaks = 1:27) +
  labs(title = "Proportion of Settled Bat Star larvae per Treatment", 
       x = NULL,
       y = "Proportion",
       color = "Treatment") +
  theme_bw() +
  theme(legend.position = "bottom")+
  theme(legend.text = element_text(size = 5))
```

Figure 2: This is a box plot graph showing the number of settled stars throughout the experiment in days for each treatment. Each treatment is represented by a different color which are labeled by the legend (control, rhodomonas only, sterilized sea grapes, and non-sterilized sea grapes). The non-sterilized sea grape treatment had a much higher settlement rate compared to the control, however settlement rate was pretty similar to the treated sea grape group.
